Mendelian randomization for causal inference in Genome-wide association studies

### 2022-01-27 mr-review.Rmd
many updates

Mendelian randomization with one instrument uses a single genetic marker 
as the proxy ("instrument" or "instrumental variable") for the "exposure" in a causal inference study. When the genetic marker is a single
nucleotide polymorphism on an autosome, the cohort is partitioned into three 
classes, two homozygotes and a heterozygote, based on genotypes for that marker.

@@

Three assumptions are central to Mendelian Randomization.

The first, often termed the

>>>>

Mendelian randomization with one instrument uses a single genetic marker 
as the proxy ("instrument" or "instrumental variable") for the "exposure" and asks 
whether variation in the genotypes at the genetic marker causes variation in the 
"outcome" variable, which is typically a second complex trait. When the genetic marker 
is a single nucleotide polymorphism on an autosome, the cohort is partitioned into three 
classes, two homozygote classes and a single heterozygote class, based on genotypes 
for that marker. When the three assumptions of mendelian randomization are satisfied, 
estimation of the causal effect of the exposure on the outcome is possible.

@@

Three assumptions are central to Mendelian Randomization. Satisfying them is necessary to infer causal effects in a MR study. Careful evaluation of the assumptions is needed. Briefly, the first assumption states that the exposure is associated with the outcome. The second requires independence, conditional on the exposure and all measured and unmeasured confounders, between the genetic marker and the outcome. The last of the three assumptions states that the genetic marker is independent of all (measured and unmeasured) confounders of the relationship between exposure and outcome. We next discuss these in more detail and consider their evaluation in MR studies.

The first assumption, about the association between the genetic marker and the exposure, holds, but this association is weak, then an amplification of biases, such as those due to violations of assumptions 2 and 3, may result [@vanderweele_methodological_2014].

The second assumption, often termed the exclusion restriction, can also be stated as 
the need for the genetic marker to affect the outcome only through the exposure. In a 
causal diagram, this means that the only path from the marker to the outcome is that 
containing the exposure.

When the three MR assumptions hold, one may use instrumental variable methods, 
with the genetic marker data, the exposure data, and the outcome data, to estimate the 
effect of the exposure on the outcome.
